---
title: "Historic Effects of Climate Change on Tree Crop Phenology in California"
author: |
  | Elise Hellwig^1^, Katherine Pope^2^, Robert J. Hijmans^1^
  | 1. University of California, Davis
  | 2. Cooperative Extension, University of California
header-includes: 
  - \usepackage{morefloats}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage[normalem]{ulem}
output:
  pdf_document:
    fig_caption: yes
  word_document:
    fig_caption: yes
    reference_docx: phenologytemplate.docx
  html_document:
    fig_caption: yes
bibliography: phenology.bib
---


```{r, env_setup, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
#This is the set-up block of code, it loads all of the required libraries

library(tidyverse)
library(knitr)
library(lubridate)
library(reshape2)
library(grid)
library(segmented)
library(kableExtra)
#library(qdap, warn.conflicts = FALSE)
library(mgsub)
library(kableExtra)

opts_chunk$set(tidy.opts=list(width.cutoff=60), 
               stringsAsFactors=FALSE,
               knitr.table.format = "latex",
               cache=TRUE,
               echo = FALSE)
options(warn=0)
#this makes it so numbers aren't put into scientific notation
#options(scipen=999)

###################################

#setwd('/Users/echellwig/Drive/Phenology/CA_Report')
#setwd("C:\\gdrive\\Phenology\\R\\CA_Report")
###################################

walnutcol <- 'green4' 
prunecol <- 'purple4'
almondcol <- 'darkgoldenrod3'

#data paths
sourcepath <- '/Users/echellwig/Research/phenology/functions'
data1 <- '/Volumes/GoogleDrive/My Drive/Phenology/CA_Results/'
data2 <- '/Volumes/GoogleDrive/My Drive/Phenology/data/historydata'
data3 <- '/Volumes/GoogleDrive/My Drive/Phenology/Results/history'
historypath <- '/Volumes/GoogleDrive/My Drive/Phenology/data/history'

source(file.path(sourcepath, 'multiplot.R'))
source(file.path(sourcepath, 'reportfunctions.R'))



#data
weather <- read.csv(file.path(data2, 'weatherlocations.csv'))
ctemps <- read.csv(file.path(data1,'switchday.csv'))
fp <- readRDS(file.path(data1, 'floweringpredictors.RDS'))
slp <- readRDS(file.path(data1, 'seasonlengthpredictors.RDS'))


am <- read.csv(file.path(data2, 'annualtemperatures.csv'))
#am <- read.csv(file.path(data1, 'data/annualtemperatures.csv'))

amlabs <- read.csv(file.path(data2, 'temptrendlabels.csv'))
names(amlabs)[2] <- 'loc'

mtc <- read.csv(file.path(data2, 'monthlytemperatures.csv'))
mtcAvg <- dcast(mtc[,c('loc', 'year', 'nmonth', 'tavg')],
                loc + year ~ nmonth, value.var = 'tavg')

harvest <- readRDS(file.path(historypath, 'harvest.RDS'))
asl <- filter(harvest, crop=='almond')
psl <- filter(harvest, crop=='prune')
wsl <- filter(harvest, crop=='walnut')

#Katherine's data

#Monthly and Total Winter Chill Accumulation for all sites (Chico, Davis, Modesto, Parlier)
wca <- read.csv(file.path(data1, 'data/monthlychillaccumulation.csv')) 
pblm <- read.csv(file.path(data1, 'data/Bloom_Parlier_French_ksp.csv')) #Prune bloom at Parlier


locations <- c('Chico', 'Modesto', 'Parlier', 'Davis')
mos <- 1:12


```



## Notes

** Thesis: Temperature increases due to climate change vary by region in California. By looking at tree response to interannual temperature variation we can explain past phenological trends and predict future ones. **

- Need to add chill and heat sums over time
- Need to run linear models and then extract the interesting numbers
- Need to find a good way to plot all the things without having massive amounts of space in the graph (ex scaling the data)
- To be honest I my want to scale the data anyway for model fitting.


# Introduction

Tree crops, like almonds and walnuts, made up 35% of the California agricultural sector by income in 2012 and that percentage is increasing (Ag stats 2012). Almonds individually are the second highest valued commodity California produces, second only to milk and cream (Ag Stats 2015). Unlike field crops and other annuals that are replanted every spring, tree crops stay in the same place for multiple decades. They also require more upfront investment as most trees do not fruit until a couple years after planting. Because of this, they are more vulnerable to changes in climate. 

The effects of climate change in California are fairly well documented, especially with respect to drought (citations). However, the potential for groundwater based irrigation systems protects many agricultural systems from significant water stress during all but the most severe droughts (Citation?). Temperatures in California are also changing due to climate change and this many already have impacted the phenology for more sensitive crops (Luedeling 2012b). Many papers have has also been demonstrated temperature to be one of the key, if not the key climatic variable when predicting crop phenology (citations). 

However, it is not enough to just look at the predictors of flowering and harvest dates when assessing climate risk. It is also necessary to compare these predictors to actually flowering and harvest dates, and investigate changes to these dates over time. 

The majority of variation in California temperatures over the past 100 years is not due to climate change, but rather is inter-annual. Since temperature variablility between years is so great, we can use crops' phenological responses over many years to fit models relating temperature and crop phenology (citation). The effects of temperature on thermal time and flowering and leaf out have also been investigated for many crops in the Central Valley using this method (Luedeling 2012). The effects on harvest however, are not as well studied.

When discussing fall phenology, the most commonly referenced development stage is maturity, the first possible harvest date [@mimoun1998; @tombesi2010; @ruml2011]. However, maturity is correlated with flowering.  Everything else being equal, a tree that bloomed earlier will also be ready to harvest earlier. Consequently, changes in harvest readiness date can be due to changes in flowering dates and/or changes in temperature after flowering dates. To avoid confounding these two phenomena, we use season length (time from flowering to maturity) as the fall phenological response, instead of harvest date.

This paper documents temperature changes in California over the past century and relates those changes to changes in almond, prune and walnut phenology in the Central Valley. Temperature increases due to climate change vary by region in California. By looking at tree response to interannual temperature variation, we can explain past phenological trends and predict future ones.

# Methods

## Data

### Climate Data

Climate data for each of the focal orchard sites was obtained from the National Climatic Data Center (NCDC, Menne and Houston 2015) and from California Irrigation Management Information System (CIMIS, California Department of Water Resources 2015). The NCDC provided daily minumum and maxiumum temperature data going back to 1906, and CIMIS provided hourly data back to 1983. Several climate stations were chosen for each orchard location to ensure there would be data for every day in the time span of the orchard datasets for each location (Table X). The closest station to the site with at least 85% completeness was chosen as the primary station (indicated by a *). Temperature data from other stations was related to temperature data from the primary station via a linear regression. This model was then used to fill in gaps in the primary station’s data. In cases where there were nearby NCDC and CIMIS weather stations, one primary station was chosen for each data source.

### 

We analyzed both the monthly and annual temperature time series with a first degree autoregressive linear model, due the presence of autocorrelation in the data. A first degree autoregressive linear model is a statistical model where the value of the dependent variable for time *t* is a predictor for the value of the dependent variable at time *t+1* [@cressie2015]. 

For plots that show data from multiple months or multiple locations, we used deviation from mean temperature, instead of just the raw temperature value, as the variable on the x-axis. So if the mean temperature value for February was 4&deg; C, points at 6&deg;C would be plotted at 2&deg; and points at 1&deg;C would be plotted at -3&deg;. This ensures that the data for any number months can be plotted together without data from the hotest or coldest month (or location) being pushed to the edge of the graph. 


```{r weatherlocs}

kable(weather, booktabs = TRUE, 
      col.names = c('Source', 'Location', 'Orchard Location','Start','End', ''))


```



### Phenological Data

Though the almond and walnut datasets have records for a large number of cultivars, only three will be analyzed – the earliest and latest cultivars to bloom and/or leaf-out, and the current most popular variety, which blooms in the middle. It is expected that if there is variability in response to warming, examining cultivars that cover the spectrum of phenological timing is the most likely way to reveal that variability. The prune industry in California is planted almost exclusively with one cultivar and thus only one cultivar will be analyzed.

Almond bloom records from the National Weather Service Fruit Frost Service (FF) and the University of California (UC) Regional Almond Variety Trial (RVT) were used. We report data for cultivars ‘Sonora’, ‘Nonpareil’, and ‘Mission’ which represent the range of bloom timing in commercially cultivated almonds.

Bloom dates reported by the National Weather Service Fruit Frost Service for the North Sacramento Valley Region, based in Chico, were derived by National Weather Service employees querying growers in the Chico area and Butte County UC Cooperative Extension.

Data from the UC Regional Almond Variety Trial were recorded in Chico and Modesto by UC Cooperative Extension advisors assigned to the county where the data was recorded. The bloom record in a given year represents the average bloom of a number of trees of similar age in the same orchard. Records are from two separate trials revisited annually. For Chico, the FF and RVT records were combined. Records spanned approximately 10-80 years, depending on the location and variety. Variety-specific data was not recorded by the FF network in Modesto, so only the more recent UCRVT data was used for Modesto.

Almond harvest records were from the UC Regional Variety Almond Trials which spanned 10 years. Similar to the almond bloom data, we report on the 'Sonora', 'Nonpareil', and 'Mission' cultivars grown in Chico, California. Prune bloom and harvest timing data was recorded by the University of California Prune Breeding, provided by current breeder Sarah Castro. Both the leaf-out and the harvest data for walnuts came from the UC Davis Walnut Improvement Program, currently headed by Charles Leslie (See Appendix).

We investigated variation and trends in the four locations in the Central Valley: Chico, Davis, Modesto, and Parlier. Temperature time series going back to 1931 were analyzed to match up with the duration of the longest phenological time series (almond bloom in Chico). While most of the phenological time series do not extend this far back, it is necessary to examine a long time series for climate data to ensure that the trends we find are robust. 


## Phenological trends


Changes in bloom or leaf-out dates over time were analyzed using both linear regression and segmented (broken-line) models. Using this dual approach can be illuminating for long-term datasets in which there are no trends or changes early in the dataset but changes do occur later in the dataset. For example, given a hundred year record of an annual event, if there has only been a change in the last 30 years, linear analysis may not pick up this change, because the consistency of the first 70 years would dilute the statistical signal of change from the last 30 years. Segmented analysis can break up this record into two trends, giving one trend line for the first 70 years and a second trend line for the 30 years after that, if this gives a more robust statistical explanation of the data.




#Results

## Temperature Trends


```{r annual_min_plot, warning=FALSE, echo=FALSE, fig.cap='Figure 3. Mean Annual Minimum Temperatures from 1930 to 2017, duration of phenology record shown with grey lines.', fig.height=5, fig.width=6, tidy=TRUE}

am_min <- am[am$variable=='tmin',]
labs_min <- amlabs[amlabs$variable=='tmin', ]

minplot <- ggplot(data=am_min) + ylim(5,13) +
    geom_line(aes(x=year, y=temp), color='blue3') + facet_wrap(~loc) +
    geom_smooth(aes(x=year, y=temp), size=1, method='lm', se=FALSE,
                color="black", formula = y ~ x) + 
     geom_segment(data=labs_min, aes(x=start, y=segy, xend=end, yend=segy),
                  color='darkgrey', size=2) + 
    geom_text(data=labs_min, aes(x=end, y=texty, label=label), hjust=1, vjust=0,
              size=3) + theme_bw(10) + 
    labs(x='Year', y=expression("Temperature  "(~degree*C)))

minplot

```


```{r annual_max_plot, warning=FALSE, echo=FALSE, fig.cap='Figure 3. Mean Annual Minimum Temperatures from 1930 to 2017, duration of phenology record shown with grey lines.', fig.height=5, fig.width=6, tidy=TRUE}

am_max <- am[am$variable=='tmax',]
labs_max <- amlabs[amlabs$variable=='tmax', ]

maxplot <- ggplot(data=am_max) + ylim(21,28) +
    geom_line(aes(x=year, y=temp), color='red3') + facet_wrap(~loc) +
    geom_smooth(aes(x=year, y=temp), size=1, method='lm', se=FALSE,
                color="black", formula = y ~ x) + 
     geom_segment(data=labs_max, aes(x=start, y=segy, xend=end, yend=segy),
                  color='darkgrey', size=2) + 
    geom_text(data=labs_max, aes(x=end, y=texty, label=label), hjust=1, vjust=0,
              size=3) + theme_bw(10) + 
    labs(x='Year', y=expression("Temperature  "(~degree*C)))

maxplot

```



Over the past 80 years, annual temperatures minimum temperatures have stayed roughly the same in Chico but the annual maximum temperature has increased by 0.085  &deg;C (Table 5, Figures 2-3). No monthly minimum temperatures in Chico exhibit any significant trends. However, January, February, and December monthly maximum temperatures show significant increasing trends (Table 5).

Modesto minimum temperatures show the strongest warming trends of all the locations, with a mean increase of 0.2 &deg;C per decade (Figure 3, Table 2). In addition, all mean minimum temperatures show significant increases in all months except for January and the means from eight months increased at over 0.3&deg;C per decade (Table 2). The annual maximum temperature in Modesto is also increasing, but not as rapidly (Figure 4, Table 5). Also, only seven months exhibit significant positive thrends in temperature (Table 5).


```{r chico_table, warning=FALSE,echo=FALSE}
chico <- filter(mtc, loc=='Chico')
achico <- am[am$loc=='Chico', ]


#running linear model and extracting coefficients and pvalues for monthly values
cmin <-  t(sapply(mos, function(m) arf('tmin', 'year', m, chico) ))
cmax <- t(sapply(mos, function(m) arf('tmax', 'year', m, chico) ))
cavg <- t(sapply(mos, function(m) arf('tavg', 'year', m, chico) ))

# #running linear model etc for annual data
cann <- t(sapply(unique(am$variable), function(v) arf('temp', 'year', NA, achico[achico$variable==v, ])))

#merging data
cmin <- rbind(cmin,cann[1,])
cmax <- rbind(cmax,cann[2,])
cavg <- rbind(cavg,cann[3,])

ctable <- data.frame(Month=c(month.name, 'Annual'), 
                     Minimum=signif(cmin[,1]*10,2), 
                     pvaln=signif(cmin[,2],2), 
                     Maximum=signif(cmax[,1]*10,2), 
                     pvalx=signif(cmax[,2],2), 
                     Average=signif(cavg[,1]*10,2),
                     pvala=signif(cavg[,2],2))

#cell_spec(Month, italic=ifelse(Month='Annual', TRUE, FALSE))

ctable %>% mutate(Month=Month,
                  Minimum = cell_spec(Minimum, bold = sig(pvaln)),
                  pvaln = cell_spec(pvaln, bold = sig(pvaln)),
                  Maximum = cell_spec(Maximum, bold = sig(pvalx)),
                  pvalx = cell_spec(pvalx, bold = sig(pvalx)),
                  Average = cell_spec(Average, bold = sig(pvala)),
                  pvala = cell_spec(pvala, bold = sig(pvala))) %>% 
    kable(digits=2, booktabs=TRUE, 
          col.names=c('Month', 'Minimum','p-value', 'Maximum', 'p-value',
                  'Average', 'p-value'), 
          caption='Average change per decade of Chico minimum, maximum, and mean monthly temperatures from 1931-2014. Change is considered significant and given in bold if p-values are under 0.05.')
    
```




Modesto minimum temperatures show the strongest warming trends of all the locations, with a mean increase of 0.2 &deg;C per decade (Figure 2, Table 6). In addition, all mean minimum temperatures show significant increases in all months except for January and the means from eight months increased at over 0.3 &deg;C per decade (Table 6). The annual maximum temperature in Modesto is also increasing, but not as rapidly (Figure 3, Table 6). Also, only seven months exhibit significant positive thrends in temperature (Table 6).



```{r modesto_table,warning=FALSE, echo=FALSE}
modesto <- filter(mtc, loc=='Modesto')
amodesto <- filter(am, loc=='Modesto')

#running AR model and extracting coefficients and pvalues

mmin <-  t(sapply(mos, function(m) arf('tmin', 'year', m, modesto) ))
mmax <- t(sapply(mos, function(m) arf('tmax', 'year', m, modesto) ))
mavg <- t(sapply(mos, function(m) arf('tavg', 'year', m, modesto) ))

mann <- t(sapply(unique(am$variable), function(v) arf('temp', 'year', NA, amodesto[amodesto$variable==v, ])))

#merging data
mmin <- rbind(mmin,mann[1,])
mmax <- rbind(mmax,mann[2,])
mavg <- rbind(mavg,mann[3,])

mtable <- data.frame(Month=c(month.name, 'Annual'), 
                     Minimum=signif(mmin[,1]*10,2), 
                     pvaln=signif(mmin[,2],2), 
                     Maximum=signif(mmax[,1]*10,2), 
                     pvalx=signif(mmax[,2],2), 
                     Average=signif(mavg[,1]*10,2),
                     pvala=signif(mavg[,2],2))

#cell_spec(Month, italic=ifelse(Month='Annual', TRUE, FALSE))

mtable %>% mutate(Month=Month,
                  Minimum = cell_spec(Minimum, bold = sig(pvaln)),
                  pvaln = cell_spec(pvaln, bold = sig(pvaln)),
                  Maximum = cell_spec(Maximum, bold = sig(pvalx)),
                  pvalx = cell_spec(pvalx, bold = sig(pvalx)),
                  Average = cell_spec(Average, bold = sig(pvala)),
                  pvala = cell_spec(pvala, bold = sig(pvala))) %>% 
    kable(digits=2, booktabs=TRUE, 
          col.names=c('Month', 'Minimum','p-value', 'Maximum', 'p-value',
                  'Average', 'p-value'), 
          caption='Average change per decade of Modesto minimum, maximum, and mean monthly temperatures from 1931-2014. Change is considered significant and given in bold if p-values are under 0.05.')

```



In Davis, annual minimum temperatures have risen an average of 0.21&deg;C per decade over the past 80 years based on an autoregressive linear analysis (Figure 2). Annual maximum temperatures, however, have stayed fairly constant (Figure 3). Davis monthly minimum temperatures exhibit strong warming trends in the spring, summer and fall, with temperatures increasing between 0.15&deg;C and 0.26&deg;C per decade in months where a significant change was detected (Table 16). January maximum temperatures in also show a significant warming trend (0.17&deg;C per decade), but it is the only month we have strong evidence of maximum temperatures increasing. In addition, mean maximum temperatures in July display a slight cooling trend (Table 7).


Davis monthly minimum temperatures exhibit strong warming trends in the spring, summer and fall, with temperatures increasing between 0.15&deg;C and 0.26&deg;C per decade in months where a significant change was detected (Table 16). January maximum temperatures in also show a significant warming trend (0.17&deg;C per decade), but it is the only month we have strong evidence of maximum temperatures increasing. In addition, mean maximum temperatures in July display a slight cooling trend (Table 16).



```{r davis_table, echo=FALSE, warning=FALSE}
davis <- mtc[mtc$loc=='Davis', ]
adavis <-  am[am$loc=='Davis', ]

#running AR model and extracting coefficients and pvalues
dmin <-  t(sapply(mos, function(m) arf('tmin', 'year', m, davis) ))
dmax <- t(sapply(mos, function(m) arf('tmax', 'year', m, davis) ))
davg <- t(sapply(mos, function(m) arf('tavg', 'year', m, davis) ))

#annual stuff
dann <-  t(sapply(unique(am$variable), function(v) arf('temp', 'year', NA, adavis[adavis$variable==v, ])))

#merging data
dmin <- rbind(dmin,dann[1,])
dmax <- rbind(dmax,dann[2,])
davg <- rbind(davg,dann[3,])


dtable <- data.frame(Month=c(month.name,'*Annual*'), Minimum=sig2(dmin[,2], v2=dmin[,1]*10), pvaln=sig2(dmin[,2]), Maximum=sig2(dmax[,2], v2=dmax[,1]*10), pvalx=sig2(dmax[,2]), Average=sig2(davg[,2], v2=davg[,1]*10), pvala=sig2(davg[,2]))

kable(dtable, digits=2, col.names=c('Month', 'Minimum','p-value', 'Maximum', 'p-value', "Average", "p-value"), caption='Average change per decade of Davis minimum, maximum, and average monthly temperatures from 1931-2014. Change is considered significant and given in bold if p-values are under 0.05.')

```


Annually Parlier exhibits warming minimum temperatures and cooling maximum temperatures (Figures 2-3) based on an autoregressive linear model of data since 1931. While the cooling trend in the mean annual maximum temperature is stronger (-0.36°C per decade), the warming trend in the mean annual minimum temperature is still highly significant (Table 8). Minimum monthly temperatures are increasing year round based on autoregression analysis, but only spring, summer, and fall maximum temperatures are cooling significantly (Table 8).

```{r parlier_table, warning=FALSE,echo=FALSE}
parlier <- mtc[mtc$loc=='Parlier', ]
aparlier <- am[am$loc=='Parlier', ]

#running AR model and extracting coefficients and pvalues
pmin <-  t(sapply(mos, function(m) arf('tmin', 'year', m, parlier) ))
pmax <- t(sapply(mos, function(m) arf('tmax', 'year', m, parlier) ))
pavg <- t(sapply(mos, function(m) arf('tavg', 'year', m, parlier) ))

pann <-  t(sapply(unique(am$variable), function(v) arf('temp', 'year', NA, aparlier[aparlier$variable==v, ])))

#merging data
pmin <- rbind(pmin,pann[1,])
pmax <- rbind(pmax,pann[2,])
pavg <- rbind(pavg,pann[3,])

mtable <- data.frame(Month=c(month.name,'*Annual*'), Minimum=sig2(pmin[,2], v2=pmin[,1]*10), pvaln=sig2(pmin[,2]), Maximum=sig2(pmax[,2], v2=pmax[,1]*10), pvalx=sig2(pmax[,2]), Averague=sig2(pavg[,2], v2=pavg[,1]*10), pvala=sig2(pavg[,2]))

kable(mtable, digits=2, col.names=c('Month', 'Minimum','p-value', 'Maximum', 'p-value','Average', 'p-value'), caption='Table 12. Average change per decade of Parlier minimum and maximum monthly temperatures from 1931-2014. Change is considered significant and given in bold if p-values are under 0.05.')

```






# Spring Phenology


##Almonds



Mean flowering dates has not changed significantly (p<0.05) for Sonora, Nonpareil, or Mission almonds over the study period in either location (Figure 5-6, Table 6). When analyzed with linear regression, the data for Nonpareil almonds in Chico showed some support for change (p=0.075); but the change was very small. The change in bloom is 0.07 days later every year (1 day every 14 years, 7 days over 100 years). The segmented analysis for Sonora almonds in Chico did detect a significant changepoint in 2006. However, this is too close to the end of the time series to be meaningful (Figure 5).


Spring bloom and heat sums were analyzed for Chico and Modesto, CA. California State University, Chico hosts the longest known bloom and harvest timing records for almonds, and is part of the long-time commercially important Northern Sacramento Valley almond growing district. Spring heat predicts flowering for both Nonpareil and Mission cultivars in Chico and Modesto. Heat sums explain more variation in Nonpareil almond bloom in both locations. January precipitation also explains some variation in Nonpareil flowering dates in Chico bringing the R$^2$ value for the final flowering model to 0.43 (Table X). 


In addition to examining chill accumulation, we analyzed when chill and heat requirements were met for almonds for the duration of the phenology records for Chico and Modesto for any potential changes. No significant changes were detected in the timing of the chill requirement being met for any of the three almond cultivars in Chico and Modesto, using either linear modeling or segmented linear modeling (Figures 13-18, Table 8-9 for p-values).  The agronomic chill requirement also did not show change over time using linear or segmented regression (Figures 13-18, Tables 8-9). 

The timing of the post-chill heat requirement being met did change significantly over time for the Sonora cultivar in both Chico and Modesto and for the Nonpareil cultivar in Modesto according to the linear models (Figures 13-18, Tables 8-9). On average the heat requirement for these cultivars was being met 0.7-0.9 days earlier every year over the course of the bloom record. . Note that in the figures below that zero is January 1st, as with Julian days. Negative numbers on the y-axis signify days in the previous calendar year. For example, Day of Year: -10 is December 22nd. Bloom data are the same as those given in Figures 5 & 6.

The fact that the heat metrics for Sonora and Nonpareil almonds have shown significant change but that there has been no change detected in the timing of bloom points to the complex inter-relatedness of a number of determinants of bloom, and requires further research to be more adequately explained. 


```{r almondbloom_time, fig.cap='Almond bloom over the past 70 years in Chico and Modesto, California.'}



aby <- ggplot(data=af) + labs(x='Year', y='Bloom Day (Julian)') +
    geom_point(aes(x=year, y=day), size = 1.5, color=almondcol) +
    facet_grid(cultivar~loc) + theme_bw() 

aby

#note there were some plots with heat sums below this but I am saving that for 
#the flowering paper.

```




```{r almondbloom_month}

afmonth <- merge(af, mtc, by=c('loc', 'year'))
afmonth <- filter(afmonth, nmonth %in% c('January'))

aftemp <- ggplot(data=afmonth) + 
    labs(x='Average January Temp', y='Bloom Day (Julian)') +
    geom_point(aes(x=tavg, y=day), size = 1.5, color=almondcol) +
    facet_grid(cultivar~loc) + theme_bw() 

aftemp

```


## Prunes


Flowering dates in Parlier increased markedly between 1988 and 2014 (Figure 35). On average, bloom has moved 0.4315 days later each year. This trend explains 41% of the variability in bloom date and was highly significant (R^2^ = 0.41, p=0.00006). 


In addition to examining chill accumulation, we analyzed when chill and heat requirements were met for prunes from 1988-2014, for any potential changes (Figure 40). No significant changes were detected in the timing of the chill requirement being met for prune (linear model p=0.69). The agronomic chill requirement also did not show change over time (Figure 40, p=0.465). The timing of the post-bloom heat requirement being met also did not change significantly over time either (Figure 40, p=0.879). 


However, prune flowering dates are getting later instead of earlier, which is contradictory to what the chill and heat accumulation, as well as the monthly temperatures imply. The fact that none of these metrics on their own explain the changing timing of bloom suggests that the influence of temperature on bloom is more complex than previously thought, or that something else entirely is influencing the flowering dates. More research is necessary to adequately explain this phenomenon.

```{r prunebloom_time, fig.cap='Improved French prune bloom over 25 years in Parlier and Chico California.', fig.height=3}

pby <- ggplot(data=pf) + labs(x='Year', y='Bloom Day (Julian)') +
    geom_point(aes(x=year, y=day), size = 1.5, color=prunecol) +
    facet_wrap(~loc) + theme_bw() 

pby

```

And some things happen in between these two plots.

```{r prunebloom_month, fig.cap="Prune flowering in Chico as a function of February Temperatures", fig.height=3, fig.width=4}

pfmonth <- merge(pf, mtc, by=c('loc', 'year'))
pfmonth <- filter(pfmonth, loc=='Chico',nmonth %in% c('February'))

#nmonth %in% c('January','February')

pftemp <- ggplot(data=pfmonth) + 
    labs(x='Average February Temp', y='Bloom Day (Julian)') +
    geom_point(aes(x=tavg, y=day), size = 1.5, color=prunecol) +
    facet_grid(~cultivar) + theme_bw() 

pftemp

```




## Walnuts


Analysis of leaf-out dates in Davis illustrate the difference in results that can be obtained if one uses linear or segmented regression with long datasets. Of the three cultivars examined (Payne, Chandler, and Franquette), only one, Chandler, showed a significant change in leaf-out timing when analyzed with a linear regression (Figure 47). However, Payne also showed low p-values when analyzed with a segmented model with a break point at 1988 (Figure 48, Table 17).

A more conceptually simple way to look at these results is by limiting the years we analyze. If we only examine the leaf-out timing for Payne and Franquette since their respective break points (1988 and 1990), we see leaf out dates getting significantly later. For 1988-2013, Payne has a p-value of 0.003 (R^2^=0.24), showing leaf-out coming later by 0.47 days/year. For data between 1990 and 2013, Franquette has a p-value of 0.015 (R^2^=0.17), showing leaf-out coming later by 0.47 days/year.





```{r Walnutbloom_time, fig.cap='Walnut leaf-out over the past 60 years in Davis, California.', fig.height=2.5}

wby <- ggplot(data=wf) + labs(x='Year', y='Leaf-out Day (Julian)') +
    geom_point(aes(x=year, y=day), size = 1.5, color=walnutcol) +
    facet_grid(~cultivar) + theme_bw() 

wby

```



```{r walnutLeaf_month, fig.cap="Walnut Leaf-out in Davis as a function of February Temperatures", fig.height=3, fig.width=6}

wf$loc <- 'Davis'
mtcm <- melt(mtc[,-3], id.vars = c('year', 'loc','nmonth'), 
             measure.vars = c('tmin','tmax', 'tavg'), value.name = 'temp')
wfmonth <- merge(wf, mtc, by=c('loc', 'year'))
wfmonth <- filter(wfmonth, loc=='Chico',nmonth %in% c('February'))

#nmonth %in% c('January','February')

pftemp <- ggplot(data=pfmonth) + 
    labs(x='Average February Temp', y='Bloom Day (Julian)') +
    geom_point(aes(x=tavg, y=day), size = 1.5, color=prunecol) +
    facet_grid(~cultivar) + theme_bw() 

pftemp

```




In addition to examining chill accumulation, we analyzed when chill and heat requirements were met for walnuts from the course of the leaf-out record (starting 1959 for Payne and Franquette, 1968 for Chandler), for any potential changes. 

For Payne, linear regression detected a significant decrease in the day of the year when the chill requirement was met (i.e. was met earlier) (p=0.04), by 0.11 days per year (Figure 55). Not surprisingly, linear regression also detected a significant change in when the agronomic chill requirement was met (p=0.04), 0.12 days earlier. Linear regression analysis did not detect a change in the timing of the heat requirement being met (p=0.59), but the segmented regression did detect change (p=0.08). According to the segmented regression, the heat requirement was met 0.22 days earlier every year until 1995. Then it was met 0.54 days later every year. This later meeting of the heat requirement may explain the later leaf-out date trend seen in Payne in recent years. 




For Chandler, linear regression did not detect a significant change in the day of the year when the chill requirement or the agronomic chill requirement was met (p=0.75 and 0.73, respectively, Figure 56). The segmented model was significant (p=0.09) for the chill requirement but results were erratic, showing a change point at 1973 and 1977, indicating these are spurious results. No change was detected in the timing of meeting the heat requirement by linear regression or segmented regression. 



For Franquette, linear regression did not detected a significant change in the day of the year when the chill requirement was met (Figure 57) by linear regression (p=0.13) or segmented regression (0.17). For the agronomic chill requirement, however, linear regression did detect a significant change in the day of the year that the agronomic requirement was met (p=0.07), coming 0.16 days earlier every year. The timing of meeting the heat requirement was not detected as changing by simple liner regression (p=0.24) or segmented regression (p=0.15). In other words, no changes in chill or heat accumulation were detected that would adequately explain the later leaf-out trends since 1990.




```{r Walnutbloom, fig.cap='Walnut leaf out by winter heat sum for three cultivars in Davis, California.', fig.height=2.5}

wb <- ggplot(data=wf) + theme_bw() + facet_wrap(~cultivar) +
    geom_point(aes(x=Heat, y=bloom), size=1, color=walnutcol) +
    labs(x='Scaled Heat Sum', y='Flowering (Julian Day)')

wb

```



```{r Walnutheatsum_time, fig.cap='Changes in walnut heat sums over the past 60 years in Davis, California.', fig.height=2.5}

wby <- ggplot(data=wf) + labs(x='Year', y='Heat Sum') +
    geom_point(aes(x=year, y=Heat), size = 1.5, color=walnutcol) +
    facet_grid(~cultivar) + theme_bw() 

wby

```




# Fall Phenology


## Almonds


Harvest dates were not available for Modesto almonds, so this section will only contain data from Chico. Mean almond season lengths appear to have decreased by varying degrees based on cultivars (Figure 23). However because the time series are so short, it is unclear whether these trends will continue into the future. 


```{r almond_harvest_year, fig.cap='Almond season lengths over 10 years in Chico, California.', warning=FALSE, echo=FALSE, fig.height=2.5}

amy <- ggplot(data=asl) + facet_wrap(~cultivar) + theme_bw() +
    geom_point(aes(x=year, y=slen), size=1.5, color=almondcol) +
    labs(x='year',y='Season Length (days)')

amy

```


```{r almond_harvest, fig.cap='Almond season length by heat sum in Chico, California.', warning=FALSE, echo=FALSE, fig.height=2.5}

amat <- ggplot(data=asl) + facet_wrap(~cultivar) + theme_bw() +
    geom_point(aes(x=heatscl, y=slen), size=1.5, color=almondcol) +
    labs(x='Scaled Heat Sum',y='Season Length (days)')

amat

```


```{r almond_gdhtime, fig.cap='Almond heat sums over 10 years in Chico, California.', warning=FALSE, echo=FALSE, fig.height=2.5}

agdh <- ggplot(data=asl) + facet_wrap(~cultivar) + theme_bw() +
    geom_point(aes(x=year, y=heatscl), size=1.5, color=almondcol) +
    labs(x='year',y='Season Length (days)')

agdh

```



## Prunes

French Prune season lengths in Parlier shortened by an average of 0.49 days per year over the past 25 years (Figure 41). Improved French prune season length was significantly correlated with average maximum May temperatures (Figure 42, Table 13). In addition, prune season length responded very strongly to fruit development thermal time accumulation calculated using both models (Figure 43, Table 14).



```{r prune_harvest_year, fig.cap='Prune season lengths over 25 years in Parlier, California.', warning=FALSE, echo=FALSE, fig.height=2.5, fig.width=4}

pmy <- ggplot(data=psl) + theme_bw() +
    geom_point(aes(x=year, y=slen), size=1.5, color=prunecol) +
    labs(x='year',y='Season Length (days)')

pmy

```


Fruit development thermal time accumulation values for French prunes in Parlier exhibit increasing trends using both models (Figure 44, Table 14). However, there is too much variation in the data to make any strong conclusions. Given the trends present in the monthly data though, it is likely that spring thermal accumulation values will continue to increase. Conversely the mean maximum May temperatures appear to be decreasing slightly (Table 12), which would indicate that prune season lengths will get longer based on the model in Figure 30. However, given that the thermal time accumulation model is more biologically accurate, it is likely to be more reliable. In this case, using monthly temperatures as an indicator of season length is misleading.

Given the trend of increasing fruit development thermal time accumulation in Parlier (Figure 44), Improved French prune season lengths will most likely continue to shrink with increasing climate change. 


```{r prune_harvest, fig.cap='Prune season lengths by heat sum in Parlier, CA.', warning=FALSE, echo=FALSE, fig.height=2.5, fig.width=4}

pmat <- ggplot(data=psl) + theme_bw() + 
    geom_point(aes(x=heatscl, y=slen), size=1.5, color=prunecol) +
    labs(x='Scaled Heat Sum',y='Season Length (days)')

pmat

```


```{r prune_heatsumtime,  fig.cap='Prune heat sum over 25 years in Parlier, CA.', warning=FALSE, echo=FALSE, fig.height=2.5, fig.width=4}

pghd <- ggplot(data=psl) + theme_bw() + 
    geom_point(aes(x=year, y=heatscl), size=1.5, color=prunecol) +
    labs(x='Year',y='Scaled Heat Sum')

pghd

```




## Walnuts


Mean season length in Payne walnuts decreased significantly over the past 50 years (Figure 58). Chandler and Franquette season lengths seem to have stayed fairly stable. However, because of the variability in the data, it is possible that a longer time series will exhibit a significant trend. This is particularly relevant for Chandler walnuts, as they were only developed in the last 50 years. 



Payne is the only walnut cultivar whose mean season length shows significant change (decrease) over the past 60 years (Figure 58). Payne season length also responded to warmer April and May temperatures (Figure 59). As monthly minimum temperatures are increasing for both April and May in Davis (Table 16), it very likely that the season length of Payne walnuts will continue to get shorter over time. Additionally, Payne season length responded strongly to fruit development thermal time accumulation during the first two months after leaf-out (Figure 60, Table 19). The fact that Payne thermal time accumulation has been increasing over the past half-century is further evidence that mean season length for Payne walnuts will shorten with increasing climate change (Figure 61, Table 19).



```{r walnut_harvest_year, fig.cap='Walnut season lengths over the past 60 years for Chandler, Franquette and Payne Cultivars.', warning=FALSE, echo=FALSE, fig.height=2.5}

wmy <- ggplot(data=wsl) + facet_wrap(~cultivar) + theme_bw() +
    geom_point(aes(x=year, y=slen), size=1.5, color=walnutcol) +
    labs(x='year',y='Season Length (days)')

wmy

```



While Chandler season length does not show a significant trend over time it does exhibit a complex response to monthly temperatures (Figure 62).  Warmer temperatures in March lead to shorter seasons, while warmer temperatures in April and May correlate with longer ones. Because March, April, and May all show warming trends in Davis (Table 16), we cannot yet tell how climate change will effect Chandler season length based on monthly temperatures. 


 
Chandler season length responds strongly to thermal time accumulation in the first thirty days after leaf-out as well (Figure 63). The trends in thermal accumulation in both models indicate that it is likely that Chandler season lengths will get shorter, even though a significant trend has not yet been detected (Figure 63-64, Table 19). It is unclear biologically why warmer temperatures in April correlate with longer season lengths, while fruit development thermal time accumulation in parts of March and April correlate with shorter season length (Figure 62-63). It may be that the correlation between April temperatures and season length is spurious.






```{r walnut_harvest, fig.cap='Heat sum correlation with walnut season lengths over the past 60 years in Davis.', warning=FALSE, echo=FALSE, fig.height=2.5}

wmat <- ggplot(data=wsl) + facet_wrap(~cultivar) + theme_bw() +
    geom_point(aes(x=heatscl, y=slen), size=1.5, color=walnutcol) +
    labs(x='Scaled Heat Sum',y='Season Length (days)')

wmat



```



While Franquette mean season length did not significantly correlate with mean temperatures from any month, it did respond to thermal time accumulation during the first 37 days after leaf-out (Figure 65, Table 19). As fruit development thermal time accumulation (calculated using the simplified model), is increasing (Figure 66, Table 19), we would expect that mean season length for Franquette walnuts will decrease as the climate warms. However, due to the amount of noise in the data, it may take a while before this trend becomes clear.



```{r walnut_heatsumtime, fig.cap='Walnut heat sums over the past 60 years.', warning=FALSE, echo=FALSE, fig.height=2.5}

wgdh <- ggplot(data=wsl) + facet_wrap(~cultivar) + theme_bw() +
    geom_point(aes(x=year, y=heatscl), size=1.5, color=walnutcol) +
    labs(x='Year',y='Scaled Heat Sum')

wgdh



```



# Conclusions

##Climate

The climate of California has not changed uniformly across the state. Modesto, Parlier, and Davis have all experienced increases in minimum annual temperatures, but the trends in their annual maximum temperatures diverge. Modesto maximum temperatures are warming, though not as quickly as the minimum temperatures. In Davis, maximum temperatures are relatively stable, and in Parlier, annual highs are actually decreasing. Additionally, both minimum and maximum temperatures in Chico are fairly stable.

## Spring Phenology

### Almond
No changes were found in the timing of bloom, nor monthly or total chill accumulation for Chico or Modesto. The only change seen was in the timing of meeting the heat requirement for Sonora and Nonpareil, the two earlier blooming cultivars, in Modesto (the warmer location) and for Nonpareil in Chico. 

It is not surprising that in Chico, where we found no change in minimum temperatures, there would be no change in chill accumulation or bloom timing. The earlier satisfaction of the heat requirement for Nonpareil in Chico is in keeping with the increased maximum temperatures in December, January and February in Chico. It is, however, difficult to explain why this was not reflected in the timing of meeting the heat requirement for the other two cultivars.

Mean monthly minimum and maximum temperatures increased every month in Modesto. It is more difficult to explain why no change was seen in bloom timing or chill accumulation in Modesto given these changes in temperature. The lack of change in chill accumulation was mirrored in the lack of change in the timing of the chill requirement being met. The increased temperatures were reflected in the earlier meeting of the heat requirement in the Nonpareil and Sonora cultivars. This may not yet have translated into a change in the timing of bloom due to the connection between chill and heat accumulation necessary for bloom.

### Prune
Bloom was found to be coming later for prunes in Parlier since 1988. Chill accumulation was found to be increasing over time for January, February and winter as a whole. This is discordant with the finding of increased mean monthly minimum temperatures in November, January, February and March, unless increased temperatures have raised some temperatures into the range giving the most chill value - between 6&deg;-8&deg;C (43&deg;-47&deg;F). Despite this change in chill accumulation, no change was found in the timing of the chill and heat requirements being met. With increased winter chill accumulation and no change in heat accumulation, one would expect bloom to be coming earlier on average, not later. More research is needed to attempt to explain if and how the change in bloom timing is related to changing temperature. 

### Walnut
For walnuts in Davis, leaf-out, our marker of spring phenology, was detected as coming later for all three cultivars since 1988. No change was found for chill accumulation in November or December, however a potential increase in chill accumulation was detected towards the end  of the century for January (since 1970) and February (1988). This change in chill accumulation does not synch with the lack of finding of significant change in minimum temperature in those months. It does, however, correspond with a finding of the chill requirement for Payne walnuts being met earlier. However, it is confusing that this was not also the finding regarding the timing of meeting the chill requirement for Chandler or Franquette. Furthermore, it is difficult to explain, in light of the increase in chill, why leaf-out has coming later in recent decades. With increased chill accumulation, one would expect leaf-out to come earlier with an earlier satisfaction of the chill requirement. More research is needed to attempt to explain if and how the change in leaf-out timing is related to changing temperature.

In short, while there is a clear signal of changing timing of spring phenology in prunes and walnuts, it is not clear if and how temperature is affecting that change.

## Fall Phenology

### Almonds
Given the length of the almond season length time series, we cannot draw any strong conclusions. However, if things continue as they have been over the ten years examined, Mission almond season lengths may potentially continue to shorten.

### Prunes

French prune season lengths have gotten, on average, 12 days shorter since 1988. The trends in the May monthly temperatures (Table 7) and the relationship between the thermal time accumulation and season length (Figure 31) both point to a continuation of this trend with continued warming due to climate change. 

### Walnuts

Payne walnut season lengths have shortened by approximately 11 days since 1960. Based on the upwards trends in the thermal time accumulation (Figure 49)  as well as April and May temperatures (Table 9), Payne season lengths will continue to get shorter as climate change intensifies. Chandler season lengths have not changed appreciably over the past 50 years. However, the increasing thermal time accumulation (Figure 52) indicates that season lengths will shorten in the future. Franquette season length shows a similar pattern to Chandler season length. It has yet to exhibit any trend, but the increasing trend in thermal accumulation for Franquette walnuts indicate that it is likely Franquette season lengths will get shorter in the future (Figure 54).



#References

