---
title: "Historic Effects of Climate Change on Tree Crop Phenology in California"
author: |
  | Elise Hellwig^1^, Katherine Pope^2^, Robert J. Hijmans^1^
  | 1. University of California, Davis
  | 2. Cooperative Extension, University of California
header-includes: 
  - \usepackage{morefloats}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage[normalem]{ulem}
output:
  pdf_document:
    fig_caption: yes
  word_document:
    fig_caption: yes
    reference_docx: phenologytemplate.docx
  html_document:
    fig_caption: yes
bibliography: phenology.bib
---


```{r, env_setup, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
#This is the set-up block of code, it loads all of the required libraries

library(tidyverse)
library(plyr)
library(knitr)
library(lubridate)
library(reshape2)
library(grid)
library(segmented)
library(kableExtra)
#library(qdap, warn.conflicts = FALSE)
library(mgsub)
library(kableExtra)

opts_chunk$set(tidy.opts=list(width.cutoff=60), 
               stringsAsFactors=FALSE,
               knitr.table.format = "latex",
               cache=TRUE,
               echo = FALSE)
options(warn=0)
#this makes it so numbers aren't put into scientific notation
#options(scipen=999)

###################################

#setwd('/Users/echellwig/Drive/Phenology/CA_Report')
#setwd("C:\\gdrive\\Phenology\\R\\CA_Report")
###################################

walnutcol <- 'green4' 
prunecol <- 'purple4'
almondcol <- 'darkgoldenrod3'

#data paths
sourcepath <- '/Users/echellwig/Research/phenology/functions'
data1 <- '/Volumes/GoogleDrive/My Drive/Phenology/CA_Results/'
data2 <- '/Volumes/GoogleDrive/My Drive/Phenology/data/historydata'
data3 <- '/Volumes/GoogleDrive/My Drive/Phenology/Results/history'
historypath <- '/Volumes/GoogleDrive/My Drive/Phenology/data/history'

source(file.path(sourcepath, 'multiplot.R'))
source(file.path(sourcepath, 'reportfunctions.R'))



#data
weather <- read.csv(file.path(data2, 'weatherlocations.csv'))
ctemps <- read.csv(file.path(data1,'switchday.csv'))
fp <- readRDS(file.path(data1, 'floweringpredictors.RDS'))
slp <- readRDS(file.path(data1, 'seasonlengthpredictors.RDS'))


am <- read.csv(file.path(data2, 'annualtemperatures.csv'))
#am <- read.csv(file.path(data1, 'data/annualtemperatures.csv'))

amlabs <- read.csv(file.path(data2, 'temptrendlabels.csv'))
names(amlabs)[2] <- 'loc'

mtc <- read.csv(file.path(data2, 'monthlytemperatures.csv'))
mtcAvg <- dcast(mtc[,c('loc', 'year', 'nmonth', 'tavg')],
                loc + year ~ nmonth, value.var = 'tavg')

harvest <- readRDS(file.path(historypath, 'harvest.RDS'))
asl <- filter(harvest, crop=='almond', cultivar!='Sonora')
psl <- filter(harvest, crop=='prune')
wsl <- filter(harvest, crop=='walnut')
tts <- readRDS(file.path(historypath, 'ThermalTimeSeries.RDS'))
locVar <- read.csv(file.path(historypath, 'SeasonLengthParameters.csv'))


#Model table
modtable <- readRDS(file.path(historypath, 'ModelTable.RDS'))
atab <- modtable %>% 
    filter(crop=='almond', var!='intercept') %>% 
    select(-c(crop,var))
ptab <- modtable %>% 
    filter(crop=='prune', var!='intercept') %>% 
    select(-c(crop,var))
wtab <- modtable %>% 
    filter(crop=='walnut', var!='intercept') %>% 
    select(-c(crop,var))

modtable$var <- recode(modtable$var, 'intercept'='intercept', .default='beta')


almondmod <- modtable %>% 
    filter(crop=='almond') %>% 
    select(-c(crop, pval, r2)) %>% 
    spread(var, coef) 

prunemod <- modtable %>% 
    filter(crop=='prune') %>% 
    select(-c(crop, pval, r2)) %>% 
    spread(var, coef) 

walnutmod <- modtable %>% 
    filter(crop=='walnut') %>% 
    select(-c(crop, pval, r2)) %>% 
    spread(var, coef) 


locations <- c('Chico', 'Modesto', 'Parlier', 'Davis')
mos <- 1:12


```


# Introduction

Fruit and nut crops, like almonds, walnuts and prunes, made up about 40% of the California agricultural sector by income in 2017 [@agstats2018]. Almonds individually are the third highest valued commodity California produces, second only to milk and cream [@agstats2018]. Unlike field crops and other annuals that are replanted every spring, tree crops stay in the same place for multiple decades. They also require more upfront investment as most trees do not fruit until a couple years after planting. Because of this, they are more vulnerable to changes in climate. 

The impact of climate change in California are fairly well documented, especially with respect to drought [@griffin2014; @diffenbaugh2015]. However, groundwater based irrigation systems currently provide a buffer for many agricultural systems from significant water stress during droughts [@medellin2016]. Temperatures in California are also changing due to climate change and this many already have impacted the phenology for more sensitive crops (Luedeling 2012b). Many papers have has also been demonstrated temperature to be one of the key, if not the key climatic variable when predicting crop phenology [@day2007;. 

However, it is not enough to just look at the predictors of flowering and harvest dates when assessing climate risk. It is also necessary to compare these predictors to actually flowering and harvest dates, and investigate changes to these dates over time. 

The majority of variation in California temperatures over the past 100 years is not due to climate change, but rather is inter-annual. Since temperature variablility between years is so great, we can use crops' phenological responses over many years to fit models relating temperature and crop phenology (citation). The effects of temperature on thermal time and flowering and leaf out have also been investigated for many crops in the Central Valley using this method (Luedeling 2012). The effects on harvest however, are not as well studied.

When discussing fall phenology, the most commonly referenced development stage is maturity, the first possible harvest date [@mimoun1998; @tombesi2010; @ruml2011]. However, maturity is correlated with flowering.  Everything else being equal, a tree that bloomed earlier will also be ready to harvest earlier. Consequently, changes in harvest readiness date can be due to changes in flowering dates and/or changes in temperature after flowering dates. To avoid confounding these two phenomena, we use season length (time from flowering to maturity) as the fall phenological response, instead of harvest date.

This paper documents temperature changes in California over the past century and relates those changes to changes in almond, prune and walnut phenology in the Central Valley. Realized temperature increases due to climate change vary by region in California. Because tree crop maturity is influenced by so many factors, detecting changes in season length over time due to climate change can be difficult. By looking at tree response to interannual temperature variation, we can explain past phenological trends and predict future ones. 


# Methods

## Data

### Climate Data

Climate data for each of the focal orchard sites was obtained from the National Climatic Data Center (NCDC, Menne and Houston 2015) and from California Irrigation Management Information System (CIMIS, California Department of Water Resources 2015). The NCDC provided daily minumum and maxiumum temperature data going back to the early 1900s, and CIMIS provided hourly data back to 1983. A number of climate stations were chosen for each orchard location to ensure there would be data for every day in the time span of the orchard datasets for each location (Table 1). The closest station to the site with at least 85% completeness was chosen as the primary station. Temperature data from other stations was related to temperature data from the primary station via a linear regression. This model was then used to fill in gaps in the primary station’s data. In cases where there were nearby NCDC and CIMIS weather stations, one primary station was chosen for each data source.

### Phenological Data

Though the almond and walnut datasets have records for a large number of cultivars, only three will be analyzed – the earliest and latest cultivars to bloom and/or leaf-out, and the current most popular variety, which blooms in the middle. It is expected that if there is variability in response to warming, examining cultivars that cover the spectrum of phenological timing is the most likely way to reveal that variability. The prune industry in California is planted almost exclusively with one cultivar and thus only one cultivar will be analyzed.

Almond bloom and harvest records came from the University of California (UC) Regional Almond Variety Trial (RAVT). Data from the UC Regional Almond Variety Trial were recorded in Chico and Modesto by UC Cooperative Extension advisors assigned to the county where the data was recorded. The bloom record in a given year represents the average bloom of a number of trees of similar age in the same orchard. We report data for cultivars ‘Sonora’, ‘Nonpareil’, and ‘Mission’ which represent the range of bloom timing in commercially cultivated almonds. Prune bloom and harvest timing data was recorded by the University of California Prune Breeding, provided by current breeder Sarah Castro. Because the 'Improved French' cultivar makes up almost all of the prune acreage in California, it is the only prune cultivar analyzed. We used walnut phenology data from the University of California at Davis Walnut Breeding Program. Leaf out dates (LD) and Harvest readiness dates (HRD) were collected by Charles Leslie, Gale McGranahan and the members of the Walnut Improvement Program. The cultivars 'Chandler', 'Payne', and 'Franquette' were chosen as cultivars that both span the range of flowering and harvest dates as well as having long data records (>30 years).


## Analysis

All analysis was completed in R Statistical Programming Environment [@r2018] using the 'phenoclim' package (citation) to estimate thermal time accumulation parameters. 

### Climate Analysis

We investigated temperature trends in the four orchard locations in the Central Valley: Chico, Davis, Modesto, and Parlier. Temperature time series going back to 1931 were analyzed to match up with the duration of the longest phenological time series (almond bloom in Chico). While most of the phenological time series do not extend this far back, it is necessary to examine a long time series for climate data to ensure that the trends we find are robust. We analyzed both the monthly and annual temperature time series with a first degree autoregressive linear model, due the presence of autocorrelation in the data [@cressie2015]. 


```{r weatherlocs}

kable(weather, booktabs = TRUE, 
      col.names = c('Source', 'Location', 'Orchard Location','Start','End', ''),
      caption = 'NCDC and CIMIS weather station information by orchard
                    location.')


```


### Phenological Analysis

Season lengths for almonds and prunes started at full bloom, while walnut season length began at leaf-out. Maturity for almonds and walnuts was counted at hull-split while prunes were considered mature at 3-4 lbs of pressure. All phenological data were analyzed separately by location and cultivar using ordinary linear regression, and segmented regression (citation) when it was deemed necessary. 

Thermal time was calculated using the asymcur model described by [@anderson1986]. We used the 'phenoclim' package to determine the optimal number of days of thermal time accumulation after bloom or leaf-out to use for predicting season length for each cultivar using 'phenoclim' package. Because latitude can effect phenological requirements (citation), we also separated data by location.

Thermal time accumulation was then calculated for each year with full bloom or leaf-out data. Next we related thermal time accumulation to season length using linear models. Finally, we examined trends in thermal time accumulation over the entire bloom time series for each cultivar and location.


#Results

## Temperature Trends


```{r annual_min_plot, warning=FALSE, echo=FALSE, fig.cap='Mean Annual Minimum Temperatures from 1930 to 2017, duration of phenology record shown with grey lines.', fig.height=5, fig.width=6, tidy=TRUE}

am_min <- am[am$variable=='tmin',]
labs_min <- amlabs[amlabs$variable=='tmin', ]

minplot <- ggplot(data=am_min) + ylim(5,13) +
    geom_line(aes(x=year, y=temp), color='blue3') + facet_wrap(~loc) +
    geom_smooth(aes(x=year, y=temp), size=1, method='lm', se=FALSE,
                color="black", formula = y ~ x) + 
     geom_segment(data=labs_min, aes(x=start, y=segy, xend=end, yend=segy),
                  color='darkgrey', size=2) + 
    geom_text(data=labs_min, aes(x=end, y=texty, label=label), hjust=1, vjust=0,
              size=3) + theme_bw(10) + 
    labs(x='Year', y=expression("Temperature  "(~degree*C)))

minplot

```


```{r annual_max_plot, warning=FALSE, echo=FALSE, fig.cap='Mean Annual Minimum Temperatures from 1930 to 2017, duration of phenology record shown with grey lines.', fig.height=5, fig.width=6, tidy=TRUE}

am_max <- am[am$variable=='tmax',]
labs_max <- amlabs[amlabs$variable=='tmax', ]

maxplot <- ggplot(data=am_max) + ylim(21,28) +
    geom_line(aes(x=year, y=temp), color='red3') + facet_wrap(~loc) +
    geom_smooth(aes(x=year, y=temp), size=1, method='lm', se=FALSE,
                color="black", formula = y ~ x) + 
     geom_segment(data=labs_max, aes(x=start, y=segy, xend=end, yend=segy),
                  color='darkgrey', size=2) + 
    geom_text(data=labs_max, aes(x=end, y=texty, label=label), hjust=1, vjust=0,
              size=3) + theme_bw(10) + 
    labs(x='Year', y=expression("Temperature  "(~degree*C)))

maxplot

```



Over the past 80 years, annual temperatures minimum temperatures have stayed roughly the same in Chico but the annual maximum temperature has increased by 0.085  &deg;C (Table 5, Figures 2-3). No monthly minimum temperatures in Chico exhibit any significant trends. However, January, February, and December monthly maximum temperatures show significant increasing trends (Table 5).

Modesto minimum temperatures show the strongest warming trends of all the locations, with a mean increase of 0.2 &deg;C per decade (Figure 3, Table 2). In addition, all mean minimum temperatures show significant increases in all months except for January and the means from eight months increased at over 0.3&deg;C per decade (Table 2). The annual maximum temperature in Modesto is also increasing, but not as rapidly (Figure 4, Table 5). Also, only seven months exhibit significant positive thrends in temperature (Table 5).


## Model Optimization

Length of optimized thermal time accumulation varied significantly between crop, location and cultivar (Table 6). Accumulation lengths varied from 35 days for Mission almonds in Modesto to 74 days for prunes in Parlier. 

```{r locVarTable}

kable(locVar[,-c(5,6)], booktabs = TRUE, 
      col.names = c('Crop', 'Location', 'Cultivar','Length'),
      caption = 'Optimized length of thermal time accumulation in days for
      predicting season length, by crop, location and cultivar.')

```


## Almonds


Mean almond season lengths did not appear to decrease over the 10 years of the Regional Almond Variety Trials. (Figure 3). However because the time series are so short, we cannot make any longterm predictions based on this. Almond season lengths respond fairly well to optimized thermal time accumulation (Figure 4, Table 7), with GDH explaining between 30% and 65% of season length variation.   Despite this, the model relating thermal time accumulation for Nonpareil Almonds to season length in Modesto was not strongly significant. 


```{r almond_harvest_year, fig.cap='Almond season lengths over 10 years in Chico, California.', warning=FALSE, echo=FALSE, fig.height=4}

amy <- ggplot(data=asl) + facet_grid(loc~cultivar) + theme_bw() +
    geom_point(aes(x=year, y=length1), size=1.5, color=almondcol) +
    labs(x='year',y='Season Length (days)')

amy

```


```{r almond_harvestthermal, fig.cap='Almond season length by heat sum in Chico, California.', warning=FALSE, echo=FALSE, fig.height=4}

AHeatLength <- filter(almondmod, predictor!='Year', response=='Season Length')

amat <- ggplot(data=asl) + facet_grid(loc~cultivar, scales='free_x') +
    geom_point(aes(x=thermal, y=length1), size=1.5, color=almondcol) +
    labs(x='Heat Sum',y='Season Length (days)') + theme_bw() +
    geom_abline(aes(intercept=intercept, slope=beta), data=AHeatLength)

amat

```



In Chico, thermal time accumulation increased by 22 GDH per year for Mission almonds and 30 GDH per year for Nonpareil almonds. In Modesto, thermal time accumulation actually decreased over time during the period of study (Table 7). This is in stark contrast to the mostly increasing mean monthly temperatures in Modesto (Table 3).


```{r almondModelTable}

kable(atab, booktabs = TRUE, 
      col.names = c('Location', 'Cultivar', 'Predictor', 'Response',
                    'Coefficient','p-value', 'R2'),
      caption = 'Summary of linear regression models relating
      year, season length and thermal time accumulation in almonds.', digits=3)

```




```{r almond_gdhtime, fig.cap='Almond heat sums over 80 years in Chico and Modesto, California.', warning=FALSE, echo=FALSE, fig.height=4}
atts <- filter(tts, crop=='almond')
AHeatYear <- filter(almondmod, predictor=='Year', response!='Season Length')

agdh <- ggplot(data=atts) + facet_grid(loc~cultivar) + theme_bw() +
    geom_point(aes(x=year, y=thermal), size=1.5, color=almondcol) +
    geom_abline(aes(intercept=intercept, slope=beta), data=AHeatYear) +
    labs(x='year',y='Heat Sum')

agdh

```



## Prunes

French Prune season lengths in Parlier shortened by an average of 0.49 days per year over the past 25 years, with a p-value of 0.041 and an adjusted R-squared of 0.160 (Figure 6, Table 8). Prune season length responded strongly to the optimized thermal time accumulation (Figure 7). Additionally, thermal time accumulated in the first 74 days after bloom showed a statistically significant increasing trend over time (Figure 8). However, this trend accounted for less variation than modeling season length over time (Table 8).



```{r prune_harvest_year, fig.cap='Prune season lengths over 25 years in Parlier, California.', warning=FALSE, echo=FALSE, fig.height=2.5, fig.width=4}

PYearLength <- filter(prunemod, predictor=='Year', response=='Season Length')


pmy <- ggplot(data=psl) + theme_bw() +
    geom_point(aes(x=year, y=length1), size=1.5, color=prunecol) +
    labs(x='year',y='Season Length (days)') +
    geom_abline(aes(intercept=intercept, slope=beta), data=PYearLength)

pmy

```



```{r prune_harvest, fig.cap='Prune season lengths by heat sum in Parlier, CA.', warning=FALSE, echo=FALSE, fig.height=2.5, fig.width=4}

PLengthHeat <- filter(prunemod, predictor=='GDH74', response=='Season Length')

pmat <- ggplot(data=psl) + theme_bw() + 
    geom_point(aes(x=thermal, y=length1), size=1.5, color=prunecol) +
    labs(x='Scaled Heat Sum',y='Season Length (days)') +
    geom_abline(aes(intercept=intercept, slope=beta), data=PLengthHeat)

pmat

```



```{r prune_heatsumtime,  fig.cap='Prune heat sum over 25 years in Parlier, CA.', warning=FALSE, echo=FALSE, fig.height=2.5, fig.width=4}

ptts <- filter(tts, crop=='prune')
PYearHeat <- filter(prunemod, predictor=='Year', response=='GDH74')


pghd <- ggplot(data=psl) + theme_bw() + 
    geom_point(aes(x=year, y=thermal), size=1.5, color=prunecol) +
    labs(x='Year',y='Heat Sum') +
    geom_abline(aes(intercept=intercept, slope=beta), data=PYearHeat)

pghd

```




```{r pruneModelTable}

kable(ptab, booktabs = TRUE, 
      col.names = c('Location', 'Cultivar', 'Predictor', 'Response',
                    'Coefficient','p-value', 'R2'),
      caption = 'Summary of linear regression models relating
      year, season length and thermal time accumulation in prunes.', digits=3)

```




## Walnuts


Mean season length in Payne walnuts decreased an average of 0.24 days per year over the past 60 years (Figure 9, Table 9). Chandler and Franquette season lengths seem to have stayed fairly stable. However, because of the variability in the data, it is possible that a longer time series will exhibit a significant trend. This is particularly relevant for Chandler walnuts, as they were only developed in the last 50 years. Season lengths for all three cultivars correlated very strongly with optimized thermal time accumulation with R^2 values ranging from 0.32 for Chandler to 0.52 for Payne (Table 9). 

Thermal time accumulation values for all three cultivars appear to be increasing over time, meaning decreasing season lengths. However, the confidence in those trends is fairly low for Chandler and Franquette cultivars. Hopefully a longer time series will either reinforce or remove those trends. 


```{r walnut_harvest_year, fig.cap='Walnut season lengths over the past 60 years for Chandler, Franquette and Payne Cultivars.', warning=FALSE, echo=FALSE, fig.height=2.5}

WYearLength <- filter(walnutmod, predictor=='Year', response=='Season Length')
WYearLength[1:2, 'beta'] <- 0


wmy <- ggplot(data=wsl) + facet_wrap(~cultivar) + theme_bw() +
    geom_point(aes(x=year, y=length1), size=1.5, color=walnutcol) +
    geom_abline(aes(intercept=intercept, slope=beta), data=WYearLength) + 
    labs(x='year',y='Season Length (days)')

wmy

```




```{r walnut_harvest, fig.cap='Heat sum correlation with walnut season lengths over the past 60 years in Davis.', warning=FALSE, echo=FALSE, fig.height=2.5}


WHeatLength <- filter(walnutmod, predictor!='Year', response=='Season Length')

wmat <- ggplot(data=wsl) + facet_wrap(~cultivar, scales='free_x') + theme_bw() +
    geom_point(aes(x=thermal, y=length1), size=1.5, color=walnutcol) +
    geom_abline(aes(intercept=intercept, slope=beta), data=WHeatLength) +
    labs(x='Scaled Heat Sum',y='Season Length (days)')

wmat



```




```{r walnut_heatsumtime, fig.cap='Walnut heat sums over the past 80 years.', warning=FALSE, echo=FALSE, fig.height=6, fig.width=5}

wtts <- filter(tts, crop=='walnut')
WHeatYear <- filter(walnutmod, predictor=='Year', response!='Season Length')
WHeatYear$signif <- c('0.1','0.1', '0.05')


wgdh <- ggplot(data=wsl) + facet_wrap(~cultivar, scale='free_y',nrow=3) +
    geom_point(aes(x=year, y=thermal), size=1.5, color=walnutcol) +
    geom_abline(aes(intercept=intercept, slope=beta, color=signif),
                data=WHeatYear) +
    scale_color_manual(values=c("#000000", "#8c8c8c"), name='') +
    labs(x='Year',y='Scaled Heat Sum') +  theme_bw() 

wgdh



```





```{r walnutModelTable}

kable(wtab, booktabs = TRUE, 
      col.names = c('Location', 'Cultivar', 'Predictor', 'Response',
                    'Coefficient','p-value', 'R2'),
      caption = 'Summary of linear regression models relating
      year, season length and thermal time accumulation in walnuts', digits=3)

```



# Discussion

##Climate

The climate of California has not changed uniformly across the state. Modesto, Parlier, and Davis have all experienced increases in minimum annual temperatures, but the trends in their annual maximum temperatures diverge. Modesto maximum temperatures are warming, though not as quickly as the minimum temperatures. In Davis, maximum temperatures are relatively stable, and in Parlier, annual highs are actually decreasing. Additionally, both minimum and maximum temperatures in Chico are fairly stable.


## Almonds

Using optimized thermal time accumulation lengths produced better predictions than previous attempts (Figure 4) [@ruml2011]. Inevitably though, a larger harvest data set would produce more robust results. Even so, despite the small number of years with harvest data for almonds, extensive almond flowering records allowed us to get a better sense of potential trends in almond season length. Trends in thermal time accumulation during the first two months after flowering point towards shortening season lengths for almonds in Chico, and lengthening ones in Modesto (Figure 5). 

## Prunes

French prune season lengths have gotten, on average, 12 days shorter between 1988 and 2014 (Figure 6). This is supported by the increasing trend in GDH74 over those same years and the strong relationship between GDH 74 and season length (Figure 7-8). We confidently expect prune season lengths will continue to shorten as the climate warms.  

## Walnuts

'Payne' walnut season lengths have shortened by approximately 11 days since 1960, while no discernable temporal trend exists yet for 'Chandler' or 'Franquette' walnuts (Figure 9). Based on the upwards trends in the thermal time accumulation (Table 5), Payne season lengths will continue to get shorter as climate change intensifies. Even though 'Chandler' and 'Franquette' season lengths have not changed appreciably over the past 50 years, there is some evidence that thermal time accumulation has (Figure 11, Table 5). Consequently, it is likely that season lengths will shorten in the future.

#Agronomic Implications

Overall, season lengths in California seem to be getting shorter. This may decrease the risk of precipitation during post-harvest drying, especially for walnuts. However, it will also likely to decrease fruit and nut size (citation). For prunes, this is problematic because fruit size strongly impacts the price farmers get for their crop. In nut crops, size is also at a premium, however there is an additional complication. Post harvest processing requires specialized equipment that is calibrated to a certain size of nut. Shorter season lengths may result in sub-optimal nut cracking, which also reduces nut quality. In many fruit crops, farmers will thin the crop to produce larger fruits. This is not common in nuts, but may become necessary as the climate warms.

#References

